---
title: "Multiple testing, summary, and final exercises"
teaching: 10
exercises: 45
questions:
- "How do we interpret p-values when mutiple comparisons are carried out?"
objectives:
- "Recognise the impact of multiple testing on the meaning of p-values"
- "Adjust p-values to compensate for the effects of multiple testing" 
keypoints:
- "Understand the impact of multiple testing on p-values"
- "Use `p.adjust` to correct multiple-testing p-values"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
gallstones <- read.csv("data/gallstones.csv", stringsAsFactors = TRUE)
gallstones[,7:12] <- data.frame(lapply(gallstones[,7:12], as.factor))
levels(gallstones$Obese) <- c("NonObese", "Obese")
levels(gallstones$Treatment) <- c("Untreated", "Treated")
levels(gallstones$Rec) <- c("NoRecurrence", "Recurrence")
levels(gallstones$Smoking.Status)<-c("NonSmoker","Smoker")
levels(gallstones$Alcohol.Consumption)<-c("NonAlcohol","Previous","Alcohol")
levels(gallstones$Mult)<-c("Single","Multiple")
library(gmodels)
library(ggplot2)
library(FSA)
happiness <- read.csv("data/happiness.csv", stringsAsFactors = TRUE)
```

## Multiple testing
In ANOVA, if H<sub>0</sub> is rejected, we carry out a _post hoc_ test to 
identify which group(s) are significantly different from the other ones. If 
there are three groups (A, B and C), that means we are performing 3 comparisons:
1. A vs B
2. A vs C
3. B vs C

Each time we perform a t-test, we chose a significance threshold under which we
reject the null hypothesis. A significance level of 0.05 for example means a 5%
chance of making a type I error - incorrectly rejecting H<sub>0</sub>. When 
performing multiple tests, that probability of error applies for each individual
comparison, so the more tests performed, the higher likelihood of a type I error.

In our example three-way comparison with signficance threshold of 0.05:
* Prob(making no mistake) = 0.95 x 0.95 x 0.95 = 0.857
* Prob(making a mistake) = 1 - 0.857 = 0.143

So rather than a 5% chance of a type I error, we instead have a much higher 
14.3% chance, which is clearly not acceptable.

> ## Challenge 1
>
> Memory test! A type I error is an incorrect rejection of the null hypothesis. 
> Can you remember what a type II error is?
> > ## Solution to Challenge 1
> > 
> > A type II error is a failure to reject the null hypothesis when it is false
> {: .solution}
{: .challenge}

### Correction for multiple testing
To compensate for the effects of multiple testing on expected error rate, we 
need to adjust the p-value according to the number of hypothesis tests 
performed. 

Examples of correction approaches include:
* False Discovery Rate (FDR): controls the type I error rate
* Bonferroni correction (FWER): reduce the probability of even one false 
discovery

In R, the `p.adjust` function can be used to perform multiple-testing correction
on a vector of p-values. 
```{r}
# Create a vector of multiple p-values to be tested
# Note, most of these *individually* meet the significance threshold of <0.05
pval <- c(0.02, 0.01, 0.05, 0.04, 0.1, 0.6, 0.5, 0.03, 0.005, 0.0005)
# False Discovery Rate correction - only three tests remain signficant
# And round to 3 decimal places for neater display
round(p.adjust(pval, method = 'fdr'),3)
# Bonferroni correction - now only one remains significant
# For some reason, bonferroni defaults to 3 decimal places anyway
p.adjust(pval, method = 'bonferroni')
```

## Summary of relevant statistical tests
This table can be used as a reference to select the appropriate test to use from
a given experimental design

|Analysis required (continuous data) |Normally distributed data |Non-normally distributed data |
|-----|-----|-----|
|Compare mean or median of one sample group against a known value |One sample t-test |Wilcoxon Rank Sum test |
|Compare means or medians of two sample groups (unpaired data) |Unpaired t-test |Mann-Whitney test |
|Compare means or medians of two sample groups (paired data) |Paired t-test |Wilcoxon Matched Pairs test |
|Compare means or medians of ≥ three sample groups (unpaired data) |ANOVA |Kruskal-Wallis ANOVA |
|Compare means or medians of ≥ three sample groups (paired data) |Repeated measures ANOVA |Friedman test |

|Analysis required (categorical data) |Test used |
|-----|-----|
|Compare two sample proportions of categorical data (unpaired data, ≤5 in one group) |Fisher exact test |
|Compare two sample proportions of categorical data (unpaired data, >5 in all groups) |Chi-square test |
|Compare two sample proportions of categorical data (paired data) |McNemar test |

## Summary of R functions

* Check for normality: `shapiro.test(var)`
* Correlation (normally distributed) `cor.test(var1,var2,method="pearson")`
* Correlation (Non-normally distributed) `cor.test(var1,var2,method="spearman")`
* Chi-square `chisq.test(table(var1,var2))`
* Fisher’s exact (≤5 counts) `fisher.test(table(var1,var2))`
* Unpaired t-test `t.test(var1~var2)`
* Paired t-test `t.test(var1~var2,paired=TRUE)`
* Wilcoxon Rank Sum test `wilcox.test(var1~var2)`
* Wilcoxon Matched Pairs Test `wilcox.test(var1~var2,paired=TRUE)`
* Compare >2 groups for one factor
* ANOVA `aov(outcome~factor1)`
* Kruskal-Wallis ANOVA `kruskal.test(outcome~factor1)`

## Final practice exercises
> ## Challenge 2
>
> Do patients with multiple gallstones experience more recurrences than those 
> with single stones?
> > ## Solution to Challenge 2
> > 
> > The variables of interest are multiple stones (categorical, 2 groups)
> > and recurrence (categorical, 2 groups). The appropriate test is therefore
> > either Chi-square or Fisher's Exact, depending on the expected count for 
> > each category
> > ```r
> > # Create the frequency table
> > table(gallstones$Rec,gallstones$Mult)
> > 
> > # Complete Cross-table
> > library(gmodels)
> > CrossTable(gallstones$Rec,gallstones$Mult,format="SPSS",prop.chisq=F,expected=T)
> > 
> > # Visualisation
> > 
> > library(ggplot2)
> > ggplot(gallstones, aes(Rec, fill=Mult)) + 
> >   geom_bar(position="fill") +
> >   scale_y_continuous(labels=scales::percent) +
> >   theme(axis.text=element_text(size=18),
> >         legend.text=element_text(size=18),
> >         legend.title=element_text(size=18),
> >         axis.title=element_text(size=18),
> >         plot.title = element_text(size=22, face="bold"))+
> >   ggtitle("Recurrence vs. Mult") 
> > 
> > # Chi-square test performed because more than 5 expected counts in each cell
> > chisq.test(gallstones$Rec,gallstones$Mult)
> > ```
> > ### Conclusion
> > With a p-value of 0.1295, there is no evidence for a significant association
> > between gallstone recurrence and the presence of multiple stones
> {: .solution}
{: .challenge}

> ## Challenge 3
>
> Is the patient's age associated with gallstone recurrence?
> > ## Solution to Challenge 3
> > 
> > The variables of interest are gallstone recurrence (categorical, 2 groups)
> > and age (continuous). The appropriate test is therefore either the t-test
> > or Mann-Whitney test, depending on whether the data is normally distributed
> > or not
> > ```r
> > # Test normality in both groups
> > shapiro.test(gallstones$Age[which(gallstones$Rec=="NoRecurrence")])
> > shapiro.test(gallstones$Age[which(gallstones$Rec=="Recurrence")])
> > 
> > # Other option
> > by(gallstones$Age,gallstones$Rec,shapiro.test)
> > 
> > # We have to perform a Mann-Whitney test 
> > wilcox.test(gallstones$Age~gallstones$Rec)
> > 
> > # Boxplots to visualise
> > plot(gallstones$Age~gallstones$Rec,col=c("red","blue"),ylab="Age",
> >      xlab="Recurrence",cex.lab=1.3)
> > 
> > # Statistical descriptions
> > by(gallstones$Age,gallstones$Rec,median)
> > by(gallstones$Age,gallstones$Rec,IQR)
> > 
> > by(gallstones$Age,gallstones$Rec,mean)
> > by(gallstones$Age,gallstones$Rec,sd)
> > ```
> > ### Conclusion
> > With a p-value of 0.7707, there is no evidence for a significant association
> > between gallstone recurrence and the age of patients
> {: .solution}
{: .challenge}

> ## Challenge 4
>
> Does alcohol consumption influence the time to gallstone dissolution?
> > ## Solution to Challenge 4
> > 
> > The variables of interest are alcohol consumption (categorical, 3 groups)
> > and dissolution time (continuous). The appropriate test is therefore either
> > ANOVA or Kruskal-Wallis, depending on whether the data is normally 
> > distributed or not
> > ```r
> > # Test normality in each groups
> > by(gallstones$Dis,gallstones$Alcohol.Consumption,shapiro.test)
> > 
> > # If distribution normal in each group
> > result<-aov(gallstones$Dis~gallstones$Alcohol.Consumption)
> > summary(result)
> > 
> > # If distribution not normal
> > kruskal.test(gallstones$Dis~gallstones$Alcohol.Consumption)
> > 
> > # Visualisation
> > plot(gallstones$Dis~gallstones$Alcohol.Consumption,col=2:4)
> > ```
> > ### Conclusion
> > With a p-value of 0.2389, there is no evidence for a significant association
> > between alcohol consumption and dissolution time
> {: .solution}
{: .challenge}

> ## Challenge 5
>
> Is there any effect of treatment and/or gender on the time to dissolution?
> > ## Solution to Challenge 5
> > 
> > The variables of interest are gender (categorical, 2 groups), treatment 
> > (categorical, two groups) and dissolution time (continuous). The appropriate
> > test is therefore a two-way ANOVA
> > ```r
> > # Visualisation
> > par(mfrow=c(1,2))
> > plot(Dis~Treatment+Gender, data=gallstones)
> > 
> > # Interaction plot to visualise
> > interaction.plot(gallstones$Treatment, gallstones$Gender, gallstones$Dis,
> >                  col=2:3,lwd=3,cex.axis=1.5,cex.lab=1.5)
> > 
> > # anova 2 way with aov function
> > result<-aov(Dis~Treatment+Gender+Treatment*Gender,data=gallstones)
> > summary(result)
> > 
> > # anova 2 way with lm function
> > result<-lm(Dis~Treatment+Gender+Treatment*Gender,data=gallstones)
> > anova(result)
> > 
> > # Checking the assumptions of anova
> > 
> > result<-lm(Dis~Treatment+Gender+Treatment*Gender,data=gallstones)
> > plot(result)
> > ```
> > ### Conclusion
> > With a p-value of <0.0016, there is evidence for a significant effect  of
> > treatment on dissolution time. There is no evidence for a difference between
> > genders for dissolution time, nor evidence for an interaction between
> > treatment and gender (both males and females have a similar response).
> {: .solution}
{: .challenge}

## Finishing up
It is always a good idea to conclude any R analysis with the `sessionInfo()`
command to save with the output as a permanent record of software versions being
used. This helps resolve issues where updated versions of tools may give 
different analysis outcomes.
```{r}
sessionInfo()
```
